---
title: "GDP Expansion-Contraction from CLI"
author: "John Balzani"
date: "11/14/2019"
output: pdf_document
---

```{r load libraries, include=FALSE}
library(tidyverse)
library(dynlm)
library(lmtest)
library(Metrics)
library(lubridate)
library(stargazer)
library(broom)
library(readr)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r import gdp data}
A191RL1Q225SBEA <- read_csv("C:/Users/John/OneDrive/Documents/Economics Study/Econometrics/GDP Prediction from 10Y3M IR Difference/A191RL1Q225SBEA.csv")
```

```{r}
OECDLOLITOAASTSAM <- read_csv("C:/Users/John/OneDrive/Documents/Economics Study/Econometrics/Recession Prediction from CLI/OECDLOLITOAASTSAM.csv")
```

```{r get qtly data}
#get 1st date of each month
CLI_filtered <- OECDLOLITOAASTSAM %>% 
  filter(DATE >= '1982-01-01') %>%
  mutate(MONTH = month(DATE)) %>% #extract month
  filter(MONTH %in% c(1, 4, 7, 10))#filter for months with qtly gdp data
```


```{r combined data}
#must have same number of rows first
combined_data_cli <- A191RL1Q225SBEA %>%
  mutate(MONTH = CLI_filtered$MONTH, CLI = CLI_filtered$OECDLOLITOAASTSAM) %>%
  mutate(CLI = CLI_filtered$OECDLOLITOAASTSAM) %>%
  mutate(real_gdp_growth = A191RL1Q225SBEA)  %>% #rename col
  mutate(gdp_impr = ifelse(real_gdp_growth > 0, 1, 0),
         mean_gdp_impr = mean(gdp_impr)) %>% #make gdp_impr & its mean
  mutate(real_gdp_growth_lag = lag(real_gdp_growth),
         real_gdp_growth_lag2 = lag(real_gdp_growth, 2),
         real_gdp_growth_lag3 = lag(real_gdp_growth, 3),
         real_gdp_growth_lag4 = lag(real_gdp_growth, 4)) %>%
  mutate(CLI_lag = lag(CLI),
         CLI_lag2 = lag(CLI, 2),
         CLI_lag3 = lag(CLI, 3),
         CLI_lag4 = lag(CLI, 4),
         CLI_lag5 = lag(CLI, 5),
         CLI_lag6 = lag(CLI, 6),
         CLI_lag7 = lag(CLI, 7),
         CLI_lag8 = lag(CLI, 8)) %>%
  mutate(delta_real_gdp_growth = real_gdp_growth - real_gdp_growth_lag,
         delta_real_gdp_growth_lag = real_gdp_growth_lag - lag(real_gdp_growth_lag),
         delta_real_gdp_growth_lag2 = real_gdp_growth_lag2 - lag(real_gdp_growth_lag2), 
         delta_real_gdp_growth_lag3 = real_gdp_growth_lag3 - lag(real_gdp_growth_lag3),
         delta_real_gdp_growth_lag4 = real_gdp_growth_lag4 - lag(real_gdp_growth_lag4)) %>%
  mutate(delta_CLI = CLI - CLI_lag,
         delta_CLI_lag = CLI_lag - CLI_lag2,
         delta_CLI_lag2 = CLI_lag2 - CLI_lag3,
         delta_CLI_lag3 = CLI_lag3 - lag(CLI_lag3),
         delta_CLI_lag4 = CLI_lag4 - lag(CLI_lag4),
         delta_CLI_lag5 = CLI_lag5 - lag(CLI_lag5),
         delta_CLI_lag6 = CLI_lag6 - lag(CLI_lag6),
         delta_CLI_lag7 = CLI_lag7 - lag(CLI_lag7),
         delta_CLI_lag8 = CLI_lag8 - lag(CLI_lag8))
  

```

```{r filter data}
combined_data_filtered <- combined_data_cli %>% filter(DATE >= "1983-01-01" & DATE < "2008-07-01") #filter to include all lags and up to 2008 recession
```

```{r plot}
combined_data_filtered %>% ggplot(aes(x = DATE)) +
  geom_line(aes(y = CLI, color = "CLI")) +
  geom_line(aes(y = real_gdp_growth, color = "real_gdp_growth")) +
  ggtitle("Figure 1: Real GDP Growth and CLI 1983-Jul 2008")
```

```{r}
combined_data_filtered_ts <- combined_data_filtered %>% ts()
```

```{r adf test gdp lag 4}
reg_gdp_adf_lags1234 <- dynlm(delta_real_gdp_growth~real_gdp_growth_lag + delta_real_gdp_growth_lag + delta_real_gdp_growth_lag2 + delta_real_gdp_growth_lag3 + delta_real_gdp_growth_lag4, data = combined_data_filtered_ts)
reg_gdp_adf_lags1234summ <- summary(reg_gdp_adf_lags1234)
reg_gdp_adf_lags1234summ
```

Conclusion: ADF test should be repeated with lag length of 3, as the absolute value of the t statistic of the last lagged value is less than 1.6.
Rule of Thumb: Set a maximum value for the lag length, and estimate the test regression with that lag length. If the the absolute value of the last lagged value in the test regression is less than 1.6, then reduce the lag length by one and retest (Ng and Perron “Lag Length Selection and the Construction of Unit Root Tests with Good Size and Power,”
ECTA, 2001.).
\newpage
ADF test with lags 1-3 of delta_real_gdp_growth:

```{r adf test gdp lag 3}
reg_gdp_adf_lags123 <- dynlm(delta_real_gdp_growth~real_gdp_growth_lag + delta_real_gdp_growth_lag + delta_real_gdp_growth_lag2 + delta_real_gdp_growth_lag3, data = combined_data_filtered_ts)
reg_gdp_adf_lags123summ <- summary(reg_gdp_adf_lags123)
reg_gdp_adf_lags123summ
```

Conclusion: The ADF test should be repeated with lag length 2, as the absolute value of the t statistic of the last lagged value is less than 1.6.
\newpage
ADF test with lags 1 and 2 of delta_real_gdp_growth:

```{r adf test gdp lag 2}
reg_gdp_adf_lags12 <- dynlm(delta_real_gdp_growth~real_gdp_growth_lag + delta_real_gdp_growth_lag + delta_real_gdp_growth_lag2, data = combined_data_filtered_ts)
reg_gdp_adf_lags12summ <- summary(reg_gdp_adf_lags12)
reg_gdp_adf_lags12summ
```

Conclusion: The ADF test should be repeated with lag length 2, as the absolute value of the t statistic of the last lagged value is less than 1.6.
\newpage
ADF test with lag 1 of delta_real_gdp_growth:

```{r adf test gdp lag 1}
reg_gdp_adf_lag1 <- dynlm(delta_real_gdp_growth~real_gdp_growth_lag + delta_real_gdp_growth_lag, data = combined_data_filtered_ts)
reg_gdp_adf_lag1summ <- summary(reg_gdp_adf_lag1)
reg_gdp_adf_lag1summ
```

Conclusion:
The t value of real_gdp_growth_lag is ```r reg_gdp_adf_lag1summ$coefficients[2, 3]```, which is below the critical value of -2.9, so we reject the null hypothesis of non-stationarity of real GDP growth. Real GDP growth is stationary.

Test for Stationarity - Augmented Dickey-Fuller Test for CLI:

Model for ADF test:
delta_CLI = alpha_adf_2 + rho1*CLI_lag + beta_adf_1*delta_CLI_lag + beta_adf_2*delta_CLI_lag_2 + beta_adf_3*delta_CLI_lag3 + beta_adf_4*delta_CLI_lag4 + epsilon_adf_2

Note: Starting with ADF test for 4 lags of delta_CLI, since it is most commonly used as a predictor of economic conditions in the following 6-9 months.
\newpage
ADF test with lags 1-4 of delta_CLI:

```{r adf test CLI 4 lags}
reg_CLI_lags1234 <- dynlm(delta_CLI~CLI_lag + delta_CLI_lag + delta_CLI_lag2 + delta_CLI_lag3 + delta_CLI_lag4, data = combined_data_filtered_ts)
reg_CLI_lags1234summ <- summary(reg_CLI_lags1234)
reg_CLI_lags1234summ
```

Conclusion: ADF test should be repeated with larger lag length, as the absolute value of the t statistic of the last lagged value is greater than 1.6.
Rule of Thumb: Set a maximum value for the lag length, and estimate the test regression with that lag length. If the the absolute value of the last lagged value in the test regression is less than 1.6, then reduce the lag length by one and retest (Ng and Perron “Lag Length Selection and the Construction of Unit Root Tests with Good Size and Power,”
ECTA, 2001.).
\newpage
ADF test with lags 1-8 of delta_CLI:

```{r adf test CLI 8 lags}
reg_CLI_8lags <- dynlm(delta_CLI~CLI_lag + delta_CLI_lag + delta_CLI_lag2 + delta_CLI_lag3 + delta_CLI_lag4 + delta_CLI_lag5 + delta_CLI_lag6 + delta_CLI_lag7 + delta_CLI_lag8, data = combined_data_filtered_ts)
reg_CLI_8lagssumm <- summary(reg_CLI_8lags)
reg_CLI_8lagssumm
```


Conclusion: The ADF test should be repeated with lag length 7, as the absolute value of the t statistic of the last lagged value is less than 1.6.
\newpage
ADF test with lags 1-7 of delta_CLI:

```{r adf test CLI 7 lags}
reg_CLI_7lags <- dynlm(delta_CLI~CLI_lag + delta_CLI_lag + delta_CLI_lag2 + delta_CLI_lag3 + delta_CLI_lag4 + delta_CLI_lag5 + delta_CLI_lag6 + delta_CLI_lag7, data = combined_data_filtered_ts)
reg_CLI_7lagssumm <- summary(reg_CLI_7lags)
reg_CLI_7lagssumm
```

Conclusion: The ADF test should be repeated with lag length 6, as the absolute value of the t statistic of the last lagged value is less than 1.6.
\newpage
ADF test with lags 1-6 of delta_CLI:

```{r adf test CLI 6 lags}
reg_CLI_6lags <- dynlm(delta_CLI~CLI_lag + delta_CLI_lag + delta_CLI_lag2 + delta_CLI_lag3 + delta_CLI_lag4 + delta_CLI_lag5 + delta_CLI_lag6, data = combined_data_filtered_ts)
reg_CLI_6lagssumm <- summary(reg_CLI_6lags)
reg_CLI_6lagssumm
```

Conclusion: The ADF test should be repeated with lag length 5, as the absolute value of the t statistic of the last lagged value is less than 1.6.
\newpage
ADF test with lags 1-5 of delta_CLI:

```{r adf test CLI 5 lags}
reg_CLI_5lags <- dynlm(delta_CLI~CLI_lag + delta_CLI_lag + delta_CLI_lag2 + delta_CLI_lag3 + delta_CLI_lag4 + delta_CLI_lag5, data = combined_data_filtered_ts)
reg_CLI_5lagssumm <- summary(reg_CLI_5lags)
reg_CLI_5lagssumm
```

The t stat of CLI_lag is ```r reg_CLI_5lagssumm$coefficients[2,3]```, which is below the critical value of -2.9, so we reject the null hypothesis of non-stationarity. CLI is stationary.

Granger Causality:
\newline
Granger Causality of Real GDP Growth:

```{r unrestricted model}
reg_ur <- dynlm(real_gdp_growth~real_gdp_growth_lag + real_gdp_growth_lag2 + real_gdp_growth_lag3 + CLI_lag + CLI_lag2 + CLI_lag3, data = combined_data_filtered_ts)
reg_ur_summ <- summary(reg_ur)
```

```{r restricted model}
reg_r <- dynlm(real_gdp_growth~real_gdp_growth_lag + real_gdp_growth_lag2 + real_gdp_growth_lag3, data = combined_data_filtered_ts)
reg_r_summ <- summary(reg_r)
```

```{r F test}
R1sqd <- reg_ur_summ$r.squared
R0sqd <- reg_r_summ$r.squared
g <- 3
n <- nrow(combined_data_filtered_ts)
k <- 7
F_stat <- ((R1sqd - R0sqd)/g)/((1-R1sqd)/(n-k))
```
\newpage
p value of F statistic:

```{r}
pf(F_stat, g, n-k, lower.tail = FALSE)
```

Conclusion: This is below 0.05, so we reject the null hypothesis of no Granger causality. The level of the CLI Granger-causes real GDP growth. If the level of the CLI Granger-causes real GDP growth, then the level of the CLI may also Granger-cause improvement or contraction of the economy.

Granger Causality of gdp_impr, a binary {0,1} variable where 0 indicates economic contraction and 1 indicates economic growth:

```{r unrestricted model}
reg_ur2 <- dynlm(gdp_impr~real_gdp_growth_lag + real_gdp_growth_lag2 + real_gdp_growth_lag3 + CLI_lag + CLI_lag2 + CLI_lag3, data = combined_data_filtered_ts)
reg_ur2summ <- summary(reg_ur)
```

```{r restricted model}
reg_r2 <- dynlm(real_gdp_growth~real_gdp_growth_lag + real_gdp_growth_lag2 + real_gdp_growth_lag3, data = combined_data_filtered_ts)
reg_r2summ <- summary(reg_r)
```

```{r F test}
R1sqd2 <- reg_ur2summ$r.squared
R0sqd2 <- reg_r2summ$r.squared
g <- 3
n <- nrow(combined_data_filtered_ts)
k <- 7
F_stat <- ((R1sqd2 - R0sqd2)/g)/((1-R1sqd2)/(n-k))
```
\newpage
p value of F statistic:

```{r}
round(pf(F_stat, g, n-k, lower.tail = FALSE), 4)
```

Conclusion: This is below 0.05, so we reject the null hypothesis of no Granger causality. The level of the CLI Granger-causes gdp_impr.

```{r logit models}
logit_model_constant <- glm(formula = gdp_impr~mean_gdp_impr, family = binomial(link = "logit"), data = combined_data_filtered)

logit_model_lags123 <- glm(formula = gdp_impr~CLI_lag + CLI_lag2 + CLI_lag3, family = binomial(link = "logit"), data = combined_data_filtered)

logit_model_lags12 <- glm(formula = gdp_impr~CLI_lag + CLI_lag2, family = binomial(link = "logit"), data = combined_data_filtered)

logit_model_lags23 <- glm(formula = gdp_impr~CLI_lag2 + CLI_lag3, family = binomial(link = "logit"), data = combined_data_filtered)

logit_model_lag1 <- glm(formula = gdp_impr~CLI_lag + CLI_lag3, family = binomial(link = "logit"), data = combined_data_filtered)

logit_model_lag2 <- glm(formula = gdp_impr~CLI_lag2, family = binomial(link = "logit"), data = combined_data_filtered)

logit_model_lag3 <- glm(formula = gdp_impr~CLI_lag3, family = binomial(link = "logit"), data = combined_data_filtered)
```


```{r Log Likelihood values}
loglike_constant <- logLik(logit_model_constant)
loglike_lags123 <- logLik(logit_model_lags123)
loglike_lags12 <- logLik(logit_model_lags12)
loglike_lags23 <- logLik(logit_model_lags23)
loglike_lag1 <- logLik(logit_model_lag1)
loglike_lags2 <- logLik(logit_model_lag2)
loglike_lags3 <- logLik(logit_model_lag3)
```

```{r Likelihood Ratio test}
LR_stat_lags123 <- lrtest(logit_model_constant, logit_model_lags123) #(restricted, unrestricted) to prove the joint significance of lags 1,2, and 3
#to prove joint significance of lags 1 and 2
LR_stat_lags12 <- lrtest(logit_model_lags12, logit_model_lags123)
#to prove joint significance of lags 2 and 3
LR_stat_lags23 <- lrtest(logit_model_lags23, logit_model_lags123)
#to prove significance of lag 1
LR_stat_lag1 <- lrtest(logit_model_lag1, logit_model_lags123)
#to prove significance of lag 2
LR_stat_lag2 <- lrtest(logit_model_lag2, logit_model_lags123)
#to prove significance of lag 3
LR_stat_lag3 <- lrtest(logit_model_lag3, logit_model_lags123)
```
    
Likelihood Ratio Statistics:
Null hypothesis H0: not all indicators are significant
Alternate hypothesis H1: all indicators significant

LR Statistic for model with lags 1,2, and 3:

```{r}
LR_stat_lags123
```

Conclusion: This model is significant at the 1% level.

LR Statistic for model with lags 1 and 2:

```{r}
LR_stat_lags12
```

Conclusion: This model is not significant at the 5% level.


LR Statistic for model with lags 2 and 3:

```{r}
LR_stat_lags23
```

Conclusion: This model is significant at the 5% level.

LR Statistic for model with lag 1:

```{r}
LR_stat_lag1
```

Conclusion: This model is not significant at the 5% level.

LR Statistic for model with lag 2:

```{r}
LR_stat_lag2
```

Conclusion: This model is significant at the 1% level.

LR Statistic for model with lag 3:

```{r}
LR_stat_lag3
```

Conclusion: This model is significant at the 1% level.

Model Comparison:

McFadden R Squared Values for Significant Models:

Model with lags 1, 2, and 3:

```{r}
mcfRsqd_lags123 <- 1 - logit_model_lags123$deviance/logit_model_lags123$null.deviance
round(mcfRsqd_lags123, 4)
```

Model with lags 2 and 3:

```{r}
mcfRsqd_lags23 <- 1 - logit_model_lags23$deviance/logit_model_lags23$null.deviance
round(mcfRsqd_lags23, 4)
```

Model with lag 2:

```{r}
mcfRsqd_lag2 <- 1 - logit_model_lag2$deviance/logit_model_lag2$null.deviance
round(mcfRsqd_lag2, 4)
```

Model with lag 3:

```{r}
mcfRsqd_lag3 <- 1 - logit_model_lag3$deviance/logit_model_lag3$null.deviance
round(mcfRsqd_lag3, 4)
```

The model with lags 1, 2, and 3 is the optimal model as it has the highest McFadden R squared value.

Predictions of Contraction or Growth:

```{r filter data for evaulation sample}
Case_GDP_round2_tbl_eval <- Case_GDP_round2_tbl %>% filter(Date >= "2011-01-01")
```